Introduction

### 2021-07-20 code/figure3.Rmd

fig5A <- ggplot(data = effect_heatmap, aes(x = cluster, y = path_name)) +
  geom_tile(size = 1, aes(fill = effect_size)) +
  scale_fill_fermenter( ## custom color scale
    type = "div",
    palette = "PuOr",
    n.breaks = 12,
    limits = c(-1, 1), ## Set scale range
    name = "Effect Size",
    labels = c(-1, "", 0.5, "", "", 0, "", "", 0.5, "", 1)
  ) +
  #facet_wrap(cluster~., ncol = 1, strip.position = "right", scales = "free_y") + ## facet by cluster
  facet_grid(.~cluster, scales = "free", space = "free", switch = "x") +
  guides(fill = guide_colorsteps(title.position = "top", title.hjust = 0.5)) +
  ylab("377 pathways, ward.D2 clustered") + xlab("Clusters") + 
  scale_x_discrete(position = "bottom") + ## move x labels to top
  guides(y.sec = guide_axis_label_trans(~ paste(.x))) + ## Duplicate axis.text on right side (cluster)
  theme(
    axis.text.x = element_blank(),
    axis.text.y = element_blank(), ## turn off y-axis text
    axis.title.x = element_text(size = 10),
    axis.title.y = element_text(size = 10),
    axis.ticks = element_blank(),
    strip.text.y = element_text(size = 10, hjust = .5),
    strip.text.y.right = element_text(angle = 0), ## Fix orientation of dupped right y-axis text
    strip.background = element_blank(),
    panel.background = element_blank(),
    panel.spacing = unit(0, "lines"),
    ## legend parameters
    legend.position = "top",
    legend.key.height = unit(0.6, "lines"),
    #legend.key.width = unit(0.6, "lines"),
    legend.text = element_text(size = 7),
    legend.title = element_text(size = 7),
    legend.margin=margin(0,0,0,0),
    legend.box.margin=margin(-10,-10,-5,-10)
  )

>>>>

fig5A <- ggplot(data = effect_heatmap, aes(x = cluster, y = path_name)) +
  geom_tile(size = 1, aes(fill = effect_size)) +
  scale_fill_fermenter( ## custom color scale
    type = "div",
    palette = "PuOr",
    n.breaks = 12,
    limits = c(-1, 1), ## Set scale range
    name = "Effect Size",
    labels = c(-0.8, "", -0.4, "", 0, "", 0.4, "", 0.8)
  ) +
  #facet_wrap(cluster~., ncol = 1, strip.position = "right", scales = "free_y") + ## facet by cluster
  facet_grid(.~cluster, scales = "free", space = "free", switch = "x") +
  guides(fill = guide_colorsteps(title.position = "top", title.hjust = 0.5)) +
  ylab("377 pathways, ward.D2 clustered") + xlab("Clusters") + 
  scale_x_discrete(position = "bottom") + ## move x labels to top
  guides(y.sec = guide_axis_label_trans(~ paste(.x))) + ## Duplicate axis.text on right side (cluster)
  theme(
    axis.text.x = element_blank(),
    axis.text.y = element_blank(), ## turn off y-axis text
    axis.title.x = element_text(size = 10),
    axis.title.y = element_text(size = 10),
    axis.ticks = element_blank(),
    strip.text.y = element_text(size = 10, hjust = .5),
    strip.text.y.right = element_text(angle = 0), ## Fix orientation of dupped right y-axis text
    strip.background = element_blank(),
    panel.background = element_blank(),
    panel.spacing = unit(0, "lines"),
    ## legend parameters
    legend.position = "top",
    legend.key.height = unit(0.6, "lines"),
    #legend.key.width = unit(0.6, "lines"),
    legend.text = element_text(size = 7),
    legend.title = element_text(size = 7),
    legend.margin=margin(0,0,0,0),
    legend.box.margin=margin(-10,-10,-5,-10)
  )

### 2021-07-20 code/figure4.Rmd

tdat_all <- all_muts %>%
  filter(Hugo_Symbol %in% all_compare_plot$Hugo_Symbol) %>%
  distinct(Tumor_Sample_Barcode, Hugo_Symbol, clust_knn_char, type) %>%
  select(type, clust_knn_char, Hugo_Symbol) %>%
  table()
tdat_all <- left_join(
  as.data.frame(tdat_all),
  clinical_plus_cluster %>%
    select(Tumor_Sample_Barcode, type, clust_knn_char) %>%
    distinct() %>%
    group_by(type, clust_knn_char) %>%
    summarise(ntumors = n())) %>%
  group_by(type, clust_knn_char) %>%
  mutate(c_prop = Freq / ntumors) %>%
  mutate(c_prop = ifelse(is.na(c_prop), 0, c_prop)) %>%
  select(type, clust_knn_char, Hugo_Symbol, c_prop) %>%
  rename(cluster = clust_knn_char)
tdat_all <- all_compare_plot %>% left_join(tdat_all)
tdat_all_p <- tdat_all %>% ggplot(aes(type, Row_Index, fill = c_prop)) +
  geom_tile() +
  scale_fill_viridis_c(option = "E", "Proportion of \nTumor Type \nMutated Within Cluster", limits = c(0, 1)) +
  xlab("Tumor Type") +
  ylab("Gene Symbol") +
  scale_y_discrete(labels = Hugo_rownames) +
  facet_grid(cluster ~ ., scales = "free_y", drop = TRUE, space = "free_y") +
  theme_classic() +
  theme(
    strip.text = element_blank(),
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5, size = 8),
    axis.text.y = element_blank(),
    axis.title.y = element_blank(),
    legend.text = element_text(size = 8),
    legend.title = element_text(size = 8),
    panel.grid.major.y = element_line(colour = "#f0f0f0")
  )

>>>>

tdat_all <- all_muts %>%
  filter(Hugo_Symbol %in% all_compare_plot$Hugo_Symbol) %>%
  distinct(Tumor_Sample_Barcode, Hugo_Symbol, clust_knn_char, type) %>%
  select(type, clust_knn_char, Hugo_Symbol) %>%
  table()
tdat_all <- left_join(
  as.data.frame(tdat_all),
  clinical_plus_cluster %>%
    select(Tumor_Sample_Barcode, type, clust_knn_char) %>%
    distinct() %>%
    group_by(type, clust_knn_char) %>%
    summarise(ntumors = n())) %>%
  group_by(type, clust_knn_char) %>%
  mutate(c_prop = Freq / ntumors) %>%
  mutate(c_prop = ifelse(is.na(c_prop), 0, c_prop)) %>%
  select(type, clust_knn_char, Hugo_Symbol, c_prop) %>%
  rename(cluster = clust_knn_char)
tdat_all <- all_compare_plot %>% left_join(tdat_all)
tdat_all_p <- tdat_all %>% ggplot(aes(type, Row_Index, fill = c_prop)) +
  geom_tile() +
  scale_fill_viridis_c(option = "E", "Proportion of \nTumor Type \nMutated Within  \nCluster", limits = c(0, 1)) +
  xlab("Tumor Type") +
  ylab("Gene Symbol") +
  scale_y_discrete(labels = Hugo_rownames) +
  facet_grid(cluster ~ ., scales = "free_y", drop = TRUE, space = "free_y") +
  theme_classic() +
  theme(
    strip.text = element_blank(),
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5, size = 8),
    axis.text.y = element_blank(),
    axis.title.y = element_blank(),
    legend.text = element_text(size = 8),
    legend.title = element_text(size = 8),
    panel.grid.major.y = element_line(colour = "#f0f0f0")
  )

### 2021-07-20 text/abstract.Rmd

Cancer is an organism-level disease, impacting processes from cellular
metabolism and the microenvironment to systemic immune response. Nevertheless,
efforts to distinguish overarching mutational processes from interactions with
the cell of origin for a tumor have seen limited success, presenting a barrier
to individualized medicine. Here we present a novel, pathway-centric approach,
extracting somatic mutational profiles within and between tissues, largely
orthogonal to cell of origin, mutational burden, or stage. Known predisposition
variants are equally distributed among clusters, and largely independent of
molecular subtype. Prognosis and risk of death vary jointly by cancer type and
cluster. Analysis of metastatic tumors reveals that differences are largely
cluster-specific and complementary, implicating convergent mechanisms that
combine familiar driver genes with diverse low-frequency lesions in
tumor-promoting pathways, ultimately producing distinct molecular phenotypes.
The results shed new light on the interplay between organism-level dysfunction
and tissue-specific lesions.

>>>>

Cancer is an organism-level disease, impacting processes from cellular
metabolism and the microenvironment to systemic immune response. Nevertheless,
efforts to distinguish overarching mutational processes from interactions with
the cell of origin for a tumor have seen limited success, presenting a barrier
to individualized medicine. Here we present a pathway-centric approach,
extracting somatic mutational profiles within and between tissues, largely
orthogonal to cell of origin, mutational burden, or stage. Known predisposition
variants are equally distributed among clusters, and largely independent of
molecular subtype. Prognosis and risk of death vary jointly by cancer type and
cluster. Analysis of metastatic tumors reveals that differences are largely
cluster-specific and complementary, implicating convergent mechanisms that
combine familiar driver genes with diverse low-frequency lesions in
tumor-promoting pathways, ultimately producing distinct molecular phenotypes.
The results shed new light on the interplay between organism-level dysfunction
and tissue-specific lesions.

### 2021-07-20 text/discussion.Rmd

Our results illustrate how unique combinations of mutations in pan-cancer driver
genes with tissue-specific pathway disruptions result in common categories when
viewed at the level of the pathway knowledge-base. Top cancer driver genes
(_e.g. PIK3CA_ and _TP53_) are found in most of the clusters, in spite of the
fact that they contribute to many cluster-specific pathways. This can only be
explained by less commonly mutated genes complementing the unique combination of driver
genes in each tumor, and we speculate that many of these less commonly mutated genes are
sensitive to increased mutation frequency in different tissues. For example, the most 
frequently mutated genes (_TP53_, _PIK3CA_ and _MUC16_) in cluster 2 were also frequently 
mutated in several other clusters. Yet, pathway enrichment in cluster 2 was defined 
by less commonly mutated genes in the proteasomal degradation pathway (_e.g. PSMA3_, 
_PSMA1_ or _PSME4_). The ubiquitin proteasome system maintains cellular protein 
homeostasis by regulating protein turnover. Pathway mutations found in cluster 2 
(_e.g._ “B-catenin degradation”, “cellular response to hypoxia”, “regulation of cell cycle” 
and _etc_) covered several aspects of the cancer hallmarks described by Hanahan 
and Weinberg. These hallmarks included “sustaining proliferative signaling”, “evading 
tumor suppressors” and “resisting cell death” [@Bhattacharjee2014]. Cluster 
3 was defined by less commonly mutated genes related to ribosomal subunit. Pathway 
enrichment for cluster 3 included dysregulation in “rRNA processing in the nucleus 
and cytosol” and consequently, pathways involved in eukaryotic translation and 
nonsense mediated decay. Mutations in the ribosomal subunit have been associated 
with selective translation of mRNA to upregulate proliferation [@kampen2019]. 
Pathway enrichment in cluster 3 consist of hallmarks relating to “sustaining proliferative 
signaling”, “evading tumor suppressors”, “resisting cell death”, “deregulating 
cellular energetics”, and “genome instability and mutations”. Cluster 4 was defined 
by less commonly mutated genes related to nucleosome structure that resulted in 
pathway enrichment in DNA methylation and DNA double-strand break repair pathway. 
Cluster 4 was also enriched for “telomere maintenance”, a mechanism contributing to 
cancer progression by counteracting telomere shortening. Mutations in the 
nucleosome as found in cluster 4 have been proposed to hit all attributes of hallmarks of cancer. Cluster 
7 and 8 shared multiple enrichment in signaling pathways related to the hallmarks: 
“sustaining proliferative signaling”, “evading tumor suppressors”, “resisting cell 
death”, and “inducing angiogenesis”. These clusters were strongly defined by 
mutations in _PI3K_ or _RAS_, respectively. Taken together, our clusters highlight 
the similarities and differences at a gene and pathway level between and within 
different cancer types – uncovering the molecular mechanism in cancer. In the future,
we hope to explore the defining features of the clusters by connecting the pathway 
mutations to cancer hallmarks.

>>>>

Our results illustrate how unique combinations of mutations in pan-cancer driver
genes with tissue-specific pathway disruptions result in common categories when
viewed at the level of the pathway knowledge-base. Top cancer driver genes
(_e.g. PIK3CA_ and _TP53_) are found in most of the clusters, in spite of the
fact that they contribute to many cluster-specific pathways. This can only be
explained by less commonly mutated genes complementing the unique combination of driver
genes in each tumor, and we speculate that many of these less commonly mutated genes are
sensitive to increased mutation frequency in different tissues. For example, the most 
frequently mutated genes (_TP53_, _PIK3CA_ and _MUC16_) in cluster 2 were also frequently 
mutated in several other clusters. Yet, pathway enrichment in cluster 2 was defined 
by less commonly mutated genes in the proteasomal degradation pathway (_e.g. PSMA3_, 
_PSMA1_ or _PSME4_). The ubiquitin proteasome system maintains cellular protein 
homeostasis by regulating protein turnover. Pathway mutations found in cluster 2 
(_e.g._ “B-catenin degradation”, “cellular response to hypoxia”, “regulation of cell cycle” 
and _etc_) covered several aspects of the cancer hallmarks described by Hanahan 
and Weinberg. These hallmarks included “sustaining proliferative signaling”, “evading 
tumor suppressors” and “resisting cell death” [@Bhattacharjee2014]. Cluster 
3 was defined by less commonly mutated genes related to ribosomal subunit. Pathway 
enrichment for cluster 3 included dysregulation in “rRNA processing in the nucleus 
and cytosol” and consequently, pathways involved in eukaryotic translation and 
nonsense mediated decay. Mutations in the ribosomal subunit have been associated 
with selective translation of mRNA to upregulate proliferation [@kampen2019]. 
Pathway enrichment in cluster 3 consist of hallmarks relating to “sustaining proliferative 
signaling”, “evading tumor suppressors”, “resisting cell death”, “deregulating 
cellular energetics”, and “genome instability and mutations”. Cluster 4 was defined 
by less commonly mutated genes related to nucleosome structure that resulted in 
pathway enrichment in DNA methylation and DNA double-strand break repair pathway. 
Cluster 4 was also enriched for “telomere maintenance”, a mechanism contributing to 
cancer progression by counteracting telomere shortening. Mutations in the 
nucleosome as found in cluster 4 have been proposed to hit all attributes of hallmarks of cancer. Cluster 
7 and 8 shared multiple enrichment in signaling pathways related to the hallmarks, 
showing that some clusters share common driver mechanisms: 
“sustaining proliferative signaling”, “evading tumor suppressors”, “resisting cell 
death”, and “inducing angiogenesis”. These clusters were strongly defined by 
mutations in _PI3K_ or _RAS_, respectively, suggesting a shared etiology in the EGFR 
pathway for which these genes represent alternative intracellular signaling mechanisms. Taken together, our clusters highlight 
the similarities and differences at a gene and pathway level between and within 
different cancer types – uncovering the molecular mechanism in cancer. In the future,
we hope to explore the defining features of the clusters by connecting the pathway 
mutations to cancer hallmarks.

### 2021-07-20 text/introduction.Rmd

More recently, pan-cancer mutations and other data have been looked at as
disruptions of normal pathway activity. Pathway-centric analysis of tumors
provides additional benefits over gene-centric analysis: (1) It is significantly 
less noisy because it aggregates molecular events across multiple genes in the 
same pathway and (2) the identification of a potential casual mechanism is easier 
to interpret as genetic aberrations are linked to molecular pathways. In a perspective, 
Creixell et al. [@Creixell-pathwaynetworkanalysis] described approaches of pathway 
and network analysis applicable to next-generation sequencing data. The most common 
approach being fixed-gene enrichment analysis to identify over-represented pathways. 
Sanchez-Vega et al. [@PCAWG-PATH] carried out an exhaustive
analysis of pathway enrichment using whole genome data. Focusing on ten frequently 
altered canonical signaling pathways in cancer, they explored the mutation frequencies 
of coding regions of genes in these pathways for each cancer type. Network approaches
address the sparseness of mutations, allowing genes to be influenced by
mutations in nearest network neighbors [@hotnet2;@hornNetSigNetworkbasedDiscovery2018;
@Iranzo-Network]. These methods are powerful especially when applied to smaller
numbers of tumors. One of the challenges in the field remains that there may be
thousands of moderate effect genes that occur at such low frequency that they
are impossible to detect using positive selection theory. Some researchers have
attempted to address this challenge by applying machine learning to cancer data
to discover groups of functionally related genes as they interact with larger
pathways and networks [@kimPancancerAnalysisSomatic2018;
@mourikis-ml-pathways; @moonlight]. These studies have proven very effective at
highlighting fundamental disease phenotypes at the pathway level across cancers
with different origins at the cellular and tissue level.

>>>>

More recently, pan-cancer mutations and other data have been looked at as
disruptions of normal pathway activity. Pathway-centric analysis of tumors
provides additional benefits over gene-centric analysis: (1) It is significantly 
less noisy because it aggregates molecular events across multiple genes in the 
same pathway and (2) the identification of a potential casual mechanism is easier 
to interpret as genetic aberrations are linked to molecular pathways. In a perspective, 
Creixell et al. [@PATHNET] described approaches of pathway 
and network analysis applicable to next-generation sequencing data, the most common 
approach being fixed-gene enrichment analysis to identify over-represented pathways. 
Sanchez-Vega et al. [@PCAWG-PATH] carried out an exhaustive
analysis of pathway enrichment using whole genome data. Focusing on ten frequently 
altered canonical signaling pathways in cancer, they explored the mutation frequencies 
of coding regions of genes in these pathways for each cancer type. Network approaches
address the sparseness of mutations, allowing genes to be influenced by
mutations in nearest network neighbors [@hotnet2;@hornNetSigNetworkbasedDiscovery2018;
@Iranzo-Network]. These methods are powerful especially when applied to smaller
numbers of tumors. One of the challenges in the field remains that there may be
thousands of moderate effect genes that occur at such low frequency that they
are impossible to detect using positive selection theory. Some researchers have
attempted to address this challenge by applying machine learning to cancer data
to discover groups of functionally related genes as they interact with larger
pathways and networks [@kimPancancerAnalysisSomatic2018;
@mourikis-ml-pathways; @moonlight]. These studies have proven very effective at
highlighting fundamental disease phenotypes at the pathway level across cancers
with different origins at the cellular and tissue level.
